---
title: "Final project"
author:
  - name: Yishan Lin
  - name: Yingqi Wang
  - name: Yiming Ding
  - name: Ruiqing Cai
date: 2024-12-11
---

## 0. Authors Information

**Yishan Lin**: JHED ID: *ylin183*; Email: *ylin183\@jh.edu*\
**Yingqi Wang**: JHED ID: *ywang911*; Email: *ywang911\@jh.edu*\
**Yiming Ding**: JHED ID: *yding75*; Email: *yding75\@jh.edu*\
**Ruiqing Cai**: JHED ID: *rcai9*; Email: *rcai9\@jh.edu*

## 1. Data Preparation and Robust Analysis

::: callout-note
## Note

In this section, data *downloading, processing,* and some *initial analysis* were conducted by using the **"functional Programming"** Paradigms
:::

```{r load required packages, warning=FALSE, include=FALSE}
packages <- c("tidyverse", "haven", "table1", "purrr", "caret", "parallel", 
              "doParallel", "mice", "randomForest", "MASS", "stargazer")
for (pkg in packages) {
  if (!require(pkg, character.only = TRUE)) {
    install.packages(pkg)
    library(pkg, character.only = TRUE)
  }
}
```

### 1.1 Download and Initial Cleaning

Data are directly downloaded from the [CDC websites](https://wwwn.cdc.gov/Nchs/Data/Nhanes). *For more information see the code*

```{r loading and initial data cleaning, message=FALSE}
# Create a folder to store all the data
if (!dir.exists("data")) { # check if folder already exists
  dir.create("data")
}

# Setup function to download the data
download_and_save <- function(url, filename) {
  filepath <- file.path("data", filename)
  
  if (!file.exists(filepath)) { # Check if file already exists
    message(paste("Downloading", filename, "..."))
    dataset <- read_xpt(url)
    saveRDS(dataset, filepath)
  } else {
    message(paste(filename, "already exists. Loading from cache."))
  }
  
  readRDS(filepath) # Load the dataset
}
 
# Function to clean individual datasets
clean_dataset <- function(data, vars) {
  data %>% dplyr::select(all_of(vars))
}

# Selecting data
datasets <- list(
  demo = list(
    url = "https://wwwn.cdc.gov/Nchs/Data/Nhanes/Public/2021/DataFiles/DEMO_L.xpt",
    file = "demo_data.rds",
    vars = c("SEQN", "RIDAGEYR", "RIAGENDR", "RIDRETH3")
  ),
  diet = list(
    url = "https://wwwn.cdc.gov/Nchs/Data/Nhanes/Public/2021/DataFiles/DR1TOT_L.xpt",
    file = "diet_data.rds",
    vars = c("SEQN", "DR1TSODI", "DR1TKCAL")
  ),
  bp = list(
    url = "https://wwwn.cdc.gov/Nchs/Data/Nhanes/Public/2021/DataFiles/BPXO_L.xpt",
    file = "bp_data.rds",
    vars = c("SEQN", "BPXOSY1", "BPXOSY2", "BPXOSY3", "BPXODI1", "BPXODI2", "BPXODI3")
  ),
  body = list(
    url = "https://wwwn.cdc.gov/Nchs/Data/Nhanes/Public/2021/DataFiles/BMX_L.xpt",
    file = "body_data.rds",
    vars = c("SEQN", "BMXBMI")
  ),
  lab = list(
    url = "https://wwwn.cdc.gov/Nchs/Data/Nhanes/Public/2021/DataFiles/TCHOL_L.xpt",
    file = "lab_data.rds",
    vars = c("SEQN", "LBXTC")
  ),
  ques = list(
    url = "https://wwwn.cdc.gov/Nchs/Data/Nhanes/Public/2021/DataFiles/BPQ_L.xpt",
    file = "ques_data.rds",
    vars = c("SEQN", "BPQ150")
  )
)

# Downloading, saving, and cleaning datasets
nhanes_data <- purrr::map(datasets, function(info) {
  raw_data <- download_and_save(info$url, info$file)
  clean_dataset(raw_data, info$vars)
})

# Merging into single dataframe
merged_data <- purrr::reduce(nhanes_data, dplyr::left_join, by = "SEQN")
```

### 1.2 Process and Table 1
**Table 1:** Characteristics of Study Participants from August 2021 to August 2023
```{r processing data and table 1, warning=FALSE}
# Function to preprocess data
process_data <- function(data) {
  data %>%
    dplyr::mutate(
      RIAGENDR = dplyr::recode_factor(RIAGENDR, `1` = "Male", `2` = "Female"),
      RIDRETH3 = dplyr::recode_factor(
        RIDRETH3,
        `1` = "Mexican American",
        `2` = "Other Hispanic",
        `3` = "Non-Hispanic White",
        `4` = "Non-Hispanic Black",
        `6` = "Other"
      ),
      BPQ150 = dplyr::recode_factor(BPQ150, `1` = "Yes", `2` = "No"),
      sybp = rowMeans(dplyr::select(., BPXOSY1, BPXOSY2, BPXOSY3), na.rm = TRUE),
      dibp = rowMeans(dplyr::select(., BPXODI1, BPXODI2, BPXODI3), na.rm = TRUE)
    )
}

# Create Table 1 function
create_table1 <- function(data) {
  table1(~ RIDAGEYR + RIAGENDR + RIDRETH3 + DR1TSODI + BMXBMI + sybp + dibp + 
           BPQ150, data = data)
}

# Labeling data
set_labels_and_units <- function(data, labels, units = NULL) {
  for (var in names(labels)) {
    label(data[[var]]) <- labels[[var]]
  }
  if (!is.null(units)) {
    for (var in names(units)) {
      units(data[[var]]) <- units[[var]]
    }
  }
  return(data)
}

## Define labels and units as named lists
labels <- list(
  RIDAGEYR = "Age at Screening",
  RIAGENDR = "Gender",
  RIDRETH3 = "Race",
  BPQ150 = "High Blood Pressure Medication",
  sybp = "Systolic Blood Pressure",
  dibp = "Diastolic Blood Pressure",
  BMXBMI = "Body Mass Index"
)

units <- list(
  RIDAGEYR = "Year",
  sybp = "mmHg",
  dibp = "mmHg",
  BMXBMI = "kg/m\u00B2"
)

# Apply the data processing function
cleaned_data <- process_data(merged_data)

# Apply labels and units
cleaned_data <- set_labels_and_units(cleaned_data, labels, units)

# Generate Table 1
create_table1(cleaned_data)

# Save cleaned data
readr::write_csv(cleaned_data, "data/NHANES_cleaned_data.csv")
```

### 1.3 Robust Anlysis
A stratified Analysis by gender was conducted. 

```{r stratified analysis}
# Stratification function setup
stratified_analysis <- function(data, group_var, analysis_func) {
  # Split data into groups
  split_data <- split(data, data[[group_var]])
  
  # Apply the analysis function and add group label
  summary_list <- map(split_data, ~ {
    analysis_func(.x) %>%
      mutate(Gender = unique(.x[[group_var]])) # Add group label
  })
  
  # Combine all group summaries into a single table
  combined_summary <- bind_rows(summary_list)
  
  # Return combined table
  combined_summary
}

summarize_data <- function(data) {
  data %>%
    summarize(
      mean_age = mean(RIDAGEYR, na.rm = TRUE),
      mean_sodium = mean(DR1TSODI, na.rm = TRUE),
      mean_sbp = mean(sybp, na.rm = TRUE),
      mean_dbp = mean(dibp, na.rm = TRUE),
      n = n()
    )
}

# stratified analysis by gender
stratified_summary <- stratified_analysis(cleaned_data, "RIAGENDR", summarize_data)
knitr::kable(stratified_summary, 
      caption = "Table 2: Summary of each variables stratified by gender", 
      format = "html")
```

## 2. Statistical Analysis for Association between Sodium Intake and Blood Pressure

```{r Regression model， warning=FALSE}
# Ensure cleaned data is available
reg_data <- cleaned_data %>%
  mutate(
    BP_Category = case_when(
      sybp < 90 ~ "Low",
      sybp >= 90 & sybp <= 130 ~ "Normal",
      sybp > 130 ~ "High"
    ) %>% factor(levels = c("Low", "Normal", "High")), # Ordered levels
    BP_Binary = ifelse(sybp > 130, 1, 0) # Binary for logistic regression
  )

# Model 1: Linear Model (Sodium Intake and Blood Pressure)
model1 <- lm(sybp ~ DR1TSODI, data = reg_data)

# Model 2: Linear Model (Age, Gender, Race, Sodium Intake)
model2 <- lm(sybp ~ DR1TSODI + RIDAGEYR + RIAGENDR + RIDRETH3, data = reg_data)

# Model 3: Linear Model (Adding more covariates)
model3 <- lm(sybp ~ DR1TSODI + RIDAGEYR + RIAGENDR + RIDRETH3 + BMXBMI + LBXTC + BPQ150 + DR1TKCAL, data = reg_data)

# Model 4: Logistic Regression (Sodium Intake and Blood Pressure)
model4 <- glm(BP_Binary ~ DR1TSODI, data = reg_data, family = binomial)

# Model 5: Ordinal Logistic Regression (Sodium Intake and Blood Pressure)
model5 <- polr(BP_Category ~ DR1TSODI, data = reg_data, Hess = TRUE)

# Model 6: Ordinal Logistic Regression (Adding age, gender, race)
model6 <- polr(BP_Category ~ DR1TSODI + RIDAGEYR + RIAGENDR + RIDRETH3, data = reg_data, Hess = TRUE)

# Model 7: Ordinal Logistic Regression (Adding more covariates)
model7 <- polr(BP_Category ~ DR1TSODI + RIDAGEYR + RIAGENDR + RIDRETH3 + BMXBMI + LBXTC + BPQ150 + DR1TKCAL, data = reg_data, Hess = TRUE)

stargazer(model1, model2, model3, model4, model5, model6, model7, type = "text", title = "Regression Models Summary")
```

